---
title: "Fiddling With Biomart"
output: html_document
---

```{r}
library(biomartr)
library(stringr)

```

```{r}
organismBM()
organismFilters("H99")

```

```{r}
listGenomes() %>%
  as_tibble() %>%
  filter(str_detect(value, "neof"))
```

```{r}
getCollection(
  db = "ensembl",
  organism="GCA_000149245",
  reference = FALSE,
  path = file.path("_db_downloads", "collections")
)
```

```{r}
getCollection(
  db = "refseq",
  organism="GCF_000149245.1",
  reference = FALSE,
  path = file.path("_db_downloads", "collections")
)
```



```{r}
getKingdoms(db= "ensembl")
getMarts()
getDatasets("fungi_mart") %>%   filter(str_detect(description,"neo"))
```


```{r}
is.genome.available(db = "refseq", organism = "H99", details = TRUE)

```

```{r}
searchDatasets(mart = "fungi_mart", pattern = "h99")

listEnsemblGenomes() 
and useEnsemblGenomes()
ensembl_fungi <- useEnsemblGenomes(biomart = "plants_mart")

```


file_path <- getGenome( db= "ensembl", 
             organism = "Arabidopsis thaliana", 
             path = file.path("_ncbi_downloads","genomes"))

Ath_genome <- read_genome(file_path, format = "fasta")


# download the genome of Arabidopsis thaliana from genbank
# and store the corresponding genome file in '_ncbi_downloads/genomes'
file_path <- getGenome( db       = "genbank", 
             organism = "Arabidopsis thaliana", 
             path = file.path("_ncbi_downloads","genomes"))

Ath_genome <- read_genome(file_path, format = "fasta")


