---
title: "The RNA Centric Analysis System Report"
author: "Bora Uyar, Dilmurat Yusuf, Ricardo Wurmus, Altuna Akalin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction - single sample analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = TRUE, fig.width = 7, fig.height = 4.2)
```


```{r load_libraries, results='hide'}
library(RCAS)
library(R.utils)
library(rtracklayer)
library(GenomicFeatures)
library(stringr)
library(genomation)
library(dplyr)
library(tidyr)

source("../common_config.R")
```

**Warning**: In this vignette, due to space limitations, we demonstrate the functions of RCAS using static images. In order to see how an interactive report from RCAS looks see `RCAS::runReport()`. 

For the most up-to-date functionality, usage and installation instructions, and example outputs, see our [github repository here](https://github.com/BIMSBbioinfo/RCAS). 

# Introduction

RCAS is an automated system that provides dynamic genome annotations for custom
input files that contain transcriptomic regions. Such transcriptomic regions
could be, for instance, peak regions detected by CLIP-Seq analysis that detect
protein-RNA interactions, RNA modifications (alias the epitranscriptome),
CAGE-tag locations, or any other collection of target regions at the level of
the transcriptome.

RCAS is designed as a reporting tool for the functional analysis of RNA-binding
sites detected by high-throughput experiments. It takes as input a BED format
file containing the genomic coordinates of the RNA binding sites and a GTF file
that contains the genomic annotation features usually provided by publicly
available databases such as Ensembl and UCSC. RCAS performs overlap operations
between the genomic coordinates of the RNA binding sites and the genomic
annotation features and produces in-depth annotation summaries such as the
distribution of binding sites with respect to gene features (exons, introns,
5’/3’ UTR regions, exon-intron boundaries, promoter regions, and whole
transcripts) along with functionally enriched term for targeted genes. Moreover,
RCAS can darry out discriminative motif discovery in the query regions. The
final report of RCAS consists of high-quality dynamic figures and tables, which
are readily applicable for publications or other academic usage.

#  Data input

RCAS minimally requires as input a BED file and a GTF file.
The BED file should contain coordinates/intervals
of transcriptomic regions
which are located via transcriptomics methods such as Clip-Seq.
The GTF file should provide reference annotation.
The recommended source of GTF files is the 
[ENSEMBLE](http://www.ensembl.org/info/data/ftp/index.html) database. 

For this vignette, in order to demonstrate RCAS functionality,
we use sample BED and GTF data that are built-in the RCAS library,
which can be imported using a common R function: data().
To import custom BED and GTF files,
the user should execute two RCAS functions called importBed() and importGtf().

## Importing sample data
```{r sample_data, eval = FALSE}
data(queryRegions) #sample queryRegions in BED format()
data(gff)          #sample GFF file
```


## Importing custom data
To use importBed() and importGtf(),
the user should provide file paths to the respective
BED file and GTF file.
To reduce memory usage and time consumption,
we advise the user to set `sampleN=10000` to avoid
huge input of intervals.


```{r}
rcas_outdir=file.path(basedir, "rcas_out")
rcas_outdir %>%
  dir.create(recursive = TRUE)

release_39_gff_url = "ftp://ftp.ensemblgenomes.org/pub/fungi/release-39/gff3/fungi_basidiomycota1_collection/cryptococcus_neoformans_var_grubii_h99/Cryptococcus_neoformans_var_grubii_h99.CNA3.39.gff3.gz"

release_39_gff_url %>%
  basename %>%
  path_ext_remove %>%
  file.path(rcas_outdir,.) ->
  h99_gff_file

h99_gff_file %>%
  path_ext_set(".bed") ->
  h99_bed_file


if (!file.exists(h99_gff_file)){
  download.file(release_39_gff_url, paste0(h99_gff_file,".gz"))
  gunzip(paste0(h99_gff_file,".gz"))
}

if (!file.exists(h99_bed_file)){
  h99_gff_file %>%
    makeTxDbFromGFF %>%
    export(h99_bed_file, "BED")
}

#------------------------------
file.path(rcas_outdir,"CNAG_06125.gff") ->
  onegene_gff_file

onegene_gff_file %>%
  path_ext_set(".bed") ->
  onegene_bed_file

if (!file.exists(onegene_gff_file)){
  h99_gff_file %>%
  read_lines %>%
  str_subset("CNAG_06125|AFR98350|##") %>%
  str_subset("###", negate=TRUE) %>%
  write_lines(onegene_gff_file)
}

if (!file.exists(onegene_bed_file)){
  onegene_gff_file %>%
    makeTxDbFromGFF %>%
    export(onegene_bed_file, "BED")
}
#------------------------------
file.path(rcas_outdir,"twogene.gff") ->
  twogene_gff_file

twogene_gff_file %>%
  path_ext_set(".bed") ->
  twogene_bed_file

if (!file.exists(twogene_gff_file)){
  h99_gff_file %>%
  read_lines %>%
  str_subset("CNAG_06125|AFR98350|CNAG_09009|AFR99111|##") %>%
  str_subset("###", negate=TRUE) %>%
  write_lines(twogene_gff_file)
}

if (!file.exists(twogene_bed_file)){
  twogene_gff_file %>%
    makeTxDbFromGFF %>%
    export(twogene_bed_file, "BED")
}

test_bam = file.path(starout_2019_dir, "SRR12933729_Aligned.sortedByCoord.out.bam")

```



# Compute transcript coverage of a set of exons.
library(GenomicRanges)
bed.file = system.file("extdata/chr21.refseq.hg19.bed", 
                       package = "genomation")
gene.parts = readTranscriptFeatures(bed.file)
transcripts = split(gene.parts$exons, gene.parts$exons$name)
transcripts = transcripts[]
myMat3 = ScoreMatrixBin(target=cage, windows=transcripts[1:250], 
                    bin.num=10)
myMat3                     




```{r}
# library(GenomicRanges)
gff.file = system.file("extdata/chr21.refseq.hg19.gtf", package = "genomation")
exons = gffToGRanges(gff.file, filter = "exon")
transcripts = split(exons, exons$transcript_id)
transcripts
```

```{r}
transcript_list = c("AFR99111","AFR98350")
exons = gffToGRanges(gtf.file, filter = "exon")
transcripts = split(exons, exons$transcript_id)
transcript_sm = ScoreMatrixBin(target = test_bam, windows = transcripts[transcript_list], bin.num = 100, strand.aware=TRUE)
plotMeta(transcript_sm)#, xcoords = c(-1000, 1000))
heatMatrix(transcript_sm, col=heat.colors(10))#, xcoords = c(-1000, 1000))
```

```{r}
transcript_sm %>%
  t %>%
  as.data.frame %>%
  rename_with(~ transcript_list) %>%
  mutate(mean_coverage = rowMeans(across(where(is.numeric))),
         position_5_3=row_number()) ->
  transcript_sm_df

transcript_sm_df %>%
  ggplot(aes(x=position_5_3,y=mean_coverage)) +
    geom_line()

transcript_sm_df %>%
  select(-mean_coverage) %>%
  pivot_longer(cols=!position_5_3, names_to="transcript", values_to="depth") %>%
  ggplot(aes(x=position_5_3, y=transcript, fill=depth)) +
    geom_tile() +
    scale_fill_gradientn(colours = heat.colors(10))
```
```{r}


```



```{r}
transcript_sm_df %>%
  ggplot(aes(V1, V2)) + 
 geom_tile()

transcript_sm_df
```



```{r}
exons = gffToGRanges(gtf.file, filter = "exon")
transcripts = split(exons, exons$transcript_id)
transcript_sm = ScoreMatrixBin(target = test_bam, windows = transcripts[c("AFR99111")], bin.num = 100, strand.aware=TRUE)
plotMeta(transcript_sm)#, xcoords = c(-1000, 1000))
heatMatrix(transcript_sm)#, xcoords = c(-1000, 1000)) -> blah
```

```{r}

```


```{r}
exons = gffToGRanges(gtf.file, filter = "exon")
transcripts = split(exons, exons$transcript_id)
transcript_sm = ScoreMatrixBin(target = test_bam, windows = transcripts[c("AFR98350")], bin.num = 100, strand.aware=TRUE)
plotMeta(transcript_sm)#, xcoords = c(-1000, 1000))
heatMatrix(transcript_sm)#, xcoords = c(-1000, 1000))
```


subset(transcriptLengths(txdb, with.cds_len = TRUE), gene_id == "100287102")




```{r}
gene.parts = readTranscriptFeatures(onegene_bed_file)
transcripts = split(gene.parts$exons, gene.parts$exons$name)
transcript_sm = ScoreMatrixBin(target = test_bam, windows = transcripts, bin.num = 100, strand.aware=TRUE)


# heatMatrix(transcript_sm)#, xcoords = c(-1000, 1000))
plotMeta(transcript_sm)#, xcoords = c(-1000, 1000))
```



#### Plot by Transcripts From GFF - One Gene
```{r}
exons = gffToGRanges(twogene_gff_file, filter = "exon")
transcripts = split(exons, exons$Parent)
transcripts_sm = ScoreMatrixBin(target = test_bam, windows = transcripts, bin.num = 100, strand.aware=TRUE)
plotMeta(transcripts_sm)#, xcoords = c(-1000, 1000))
```




#### Plot by Transcripts From GFF - One Gene
```{r}
exons = gffToGRanges(onegene_gff_file, filter = "exon")
transcripts = split(exons, exons$Parent)
transcripts_sm = ScoreMatrixBin(target = test_bam, windows = transcripts, bin.num = 100, strand.aware=TRUE)
plotMeta(transcripts_sm)#, xcoords = c(-1000, 1000))
```


#### Plot by Transcripts From GFF - All Genes
```{r}
exons = gffToGRanges(h99_gff_file, filter = "exon")
transcripts = split(exons, exons$Parent)
transcripts

# transcripts_sm = ScoreMatrixBin(target = test_bam, windows = transcripts, bin.num = 100, strand.aware=TRUE)
# plotMeta(transcripts_sm)#, xcoords = c(-1000, 1000))
```

```{r}
import(onegene_gff_file)
```


```{r}
exons
```

#### Try to subset features
```{r}
exons = gffToGRanges(h99_gff_file, filter = "exon")


transcripts = split(exons, exons$Parent)
transcripts_sm = ScoreMatrixBin(target = test_bam, windows = transcripts, bin.num = 100, strand.aware=TRUE)
plotMeta(transcripts_sm)#, xcoords = c(-1000, 1000))
```




exons_sm = ScoreMatrixBin(target = test_bam, windows = exons, bin.num = 100)
plotMeta(exons_sm)#, xcoords = c(-1000, 1000))
```



```{r}
test_bam = file.path(starout_2019_dir, "SRR12933729_Aligned.sortedByCoord.out.bam")
# exons = gffToGRanges(h99_gff_file, filter = "exon")
exons = gffToGRanges(onegene_gff_file, filter = "exon")
sm = ScoreMatrixBin(target = test_bam, windows = exons, bin.num = 50)
```

```{r}
test_bam = file.path(starout_2019_dir, "SRR12933729_Aligned.sortedByCoord.out.bam")
# exons = gffToGRanges(h99_gff_file, filter = "exon")
genes = gffToGRanges(onegene_gff_file, filter = "gene")
gene_sm = ScoreMatrixBin(target = test_bam, windows = genes, bin.num = 2316, bin.op="sum")
plotMeta(gene_sm)#, xcoords = c(-1000, 1000))
```

```{r}
ScoreMatrix(target = test_bam, windows = genes) %>%
  min

```



```{r}
min(gene_sm)
```



```{r}
test_bam = file.path(starout_2019_dir, "SRR12933729_Aligned.sortedByCoord.out.bam")
# exons = gffToGRanges(h99_gff_file, filter = "exon")
transcripts = gffToGRanges(onegene_gff_file, filter = "transcript")
transcript_sm = ScoreMatrixBin(target = test_bam, windows = exons, bin.num = 50)
```


```{r}
test_bam = file.path(starout_2019_dir, "SRR12933729_Aligned.sortedByCoord.out.bam")
# exons = gffToGRanges(h99_gff_file, filter = "exon")
exons = gffToGRanges(onegene_gff_file, filter = "exon")
transcripts = split(exons, exons$transcript_id)
# sm = ScoreMatrixBin(target = test_bam, windows = exons, bin.num = 50)

sm = ScoreMatrixBin(target = test_bam, windows = transcripts, bin.num = 10)
# sm
```














```{r RCAS_import_data}
queryRegions <- importBed(filePath = <path to BED file>, sampleN = 10000)
gff <- importGtf(filePath = gtf_with_mito_rrna.file)
```

# Summarizing Overlaps of Query Regions with Genomic Annotation Features

## Querying the annotation file 
```{r queryGFF}
overlaps <- as.data.table(queryGff(queryRegions = queryRegions, gffData = gff))
```


### Finding targeted gene types
To find out the distribution of the query regions across gene types:
```{r query_gene_types}
biotype_col <- grep('gene_biotype', colnames(overlaps), value = T)
df <- overlaps[,length(unique(queryIndex)), by = biotype_col]
colnames(df) <- c("feature", "count")
df$percent <- round(df$count / length(queryRegions) * 100, 1)
df <- df[order(count, decreasing = TRUE)]
ggplot2::ggplot(df, aes(x = reorder(feature, -percent), y = percent)) + 
  geom_bar(stat = 'identity', aes(fill = feature)) + 
  geom_label(aes(y = percent + 0.5), label = df$count) + 
  labs(x = 'transcript feature', y = paste0('percent overlap (n = ', length(queryRegions), ')')) + 
  theme_bw(base_size = 14) + 
  theme(axis.text.x = element_text(angle = 90))
```

## Extending the annotation feature space 

GTF files contain some annotation features (e.g. exons, transcripts) that are 
usually explicitly defined, however, some transcript features such as introns,
exon-intron boundaries, promoter regions are only implicitly defined. Such 
implicit features can be extracted from a GTF file using makeTxDb family of 
functions from the GenomicFeatures library.

First we create a list of GRanges objects, where each list element contains all
the available coordinates of transcript features such as transcripts, exons, introns,
5'/3' UTRs, exon-intron boundaries, and promoter regions. 

```{r getTxdbFeatures}
txdbFeatures <- getTxdbFeaturesFromGRanges(gff)
```

### Plotting overlap counts between query regions and transcript features
To have a global overview of the distribution of query regions across gene
features, we can use the summarizeQueryRegions function. If a given query region 
does not overlap with any of the given coordinates of the transcript features, 
it is categorized under `NoFeatures`. 

```{R summarizeQueryRegions}
summary <- summarizeQueryRegions(queryRegions = queryRegions, 
                                 txdbFeatures = txdbFeatures)

df <- data.frame(summary)
df$percent <- round((df$count / length(queryRegions)), 3) * 100
df$feature <- rownames(df)
ggplot2::ggplot(df, aes(x = reorder(feature, -percent), y = percent)) + 
  geom_bar(stat = 'identity', aes(fill = feature)) + 
  geom_label(aes(y = percent + 3), label = df$count) + 
  labs(x = 'transcript feature', y = paste0('percent overlap (n = ', length(queryRegions), ')')) + 
  theme_bw(base_size = 14) + 
  theme(axis.text.x = element_text(angle = 90))
```

### Obtaining a table of overlap counts between query regions and genes 
To find out which genes overlap with how many queries and categorise overlaps by
transcript features; we use `getTargetedGenesTable` function, which returns a data.frame
object.  
```{r getTargetedGenesTable}
dt <- getTargetedGenesTable(queryRegions = queryRegions, 
                           txdbFeatures = txdbFeatures)
dt <- dt[order(transcripts, decreasing = TRUE)]

knitr::kable(dt[1:10,])
```

### Profiling the coverage of query regions across transcript features 

#### Coverage profile of query regions at feature boundaries 

It may be useful to look at the distribution of query regions at the boundaries of transcript features.
For instance, it may be important to see the relative signal at transcript ends (transcription start sites
versus transcription end sites). Or, it may be important to see how the signal is distributed at exon boundaries, which may give an idea about the regulation of the transcript. Here we demonstrate how to
get such signal distributions at transcription start/end sites. The same approach can be done for any other
collection of transcript features (exons, introns, promoters, UTRs etc.)

```{r transcriptBoundaryCoverage}
cvgF <- getFeatureBoundaryCoverage(queryRegions = queryRegions, 
                                   featureCoords = txdbFeatures$transcripts, 
                                   flankSize = 1000, 
                                   boundaryType = 'fiveprime', 
                                   sampleN = 10000)
cvgT <- getFeatureBoundaryCoverage(queryRegions = queryRegions, 
                                   featureCoords = txdbFeatures$transcripts, 
                                   flankSize = 1000, 
                                   boundaryType = 'threeprime', 
                                   sampleN = 10000)

cvgF$boundary <- 'fiveprime'
cvgT$boundary <- 'threeprime'

df <- rbind(cvgF, cvgT)

ggplot2::ggplot(df, aes(x = bases, y = meanCoverage)) + 
  geom_ribbon(fill = 'lightgreen', 
              aes(ymin = meanCoverage - standardError * 1.96, 
                  ymax = meanCoverage + standardError * 1.96)) + 
 geom_line(color = 'black') + 
 facet_grid( ~ boundary) + theme_bw(base_size = 14) 

```

#### Coverage profile of query regions for all transcript features
Coverage profiles can be obtained for a single type of transcript feature or a list of
transcript features. Here we demonstrate how to get coverage profile of query regions 
across all available transcript features. It might be a good idea to use sampleN
parameter to randomly downsample the target regions to speed up the 
calculations.

```{r coverageprofilelist, fig.height=6}
cvgList <- calculateCoverageProfileList(queryRegions = queryRegions, 
                                       targetRegionsList = txdbFeatures, 
                                       sampleN = 10000)

ggplot2::ggplot(cvgList, aes(x = bins, y = meanCoverage)) + 
  geom_ribbon(fill = 'lightgreen', 
              aes(ymin = meanCoverage - standardError * 1.96, 
                  ymax = meanCoverage + standardError * 1.96)) + 
 geom_line(color = 'black') + theme_bw(base_size = 14) +
 facet_wrap( ~ feature, ncol = 3) 
```

# Discriminative Motif Discovery

## Calculating enriched motifs

We build a classifier based on k-mer frequencies (allowing for mismatches) to 
find the most informative motifs that help discriminate the query sequences from the 
background distribution. 

```{r motif_analysis}
motifResults <- runMotifDiscovery(queryRegions = queryRegions, 
                           resizeN = 15, sampleN = 10000,
                           genomeVersion = 'hg19', motifWidth = 6,
                           motifN = 2, nCores = 1)

ggseqlogo::ggseqlogo(motifResults$matches_query)
```

## motif analysis: getting motif summary statistics
A summary table from the motif analysis results can be obtained
```{r motif_analysis_table}
summary <- getMotifSummaryTable(motifResults)
knitr::kable(summary)
```


# Functional enrichment  analysis

RCAS makes use of [gprofiler2](https://cran.r-project.org/web/packages/gprofiler2/index.html) package 
enriched functions in genes that overlap the query regions. 

```{r GO analysis}
targetedGenes <- unique(overlaps$gene_id)

res <- RCAS::findEnrichedFunctions(targetGenes = targetedGenes, species = 'hsapiens')
res <- res[order(res$p_value),]
resGO <- res[grep('GO:BP', res$source),]
knitr::kable(subset(resGO[1:10,], select = c('p_value', 'term_name', 'source')))
```

# Generating a full report 

The users can use the runReport() function to generate full custom reports 
including all the analysis modules described above. There are four main parts of
the analysis report.

- Annotation summaries via overlap operations 
- Functional term analysis 
- Motif analysis

By default, runReport() function aims to run all three modules, while the user
can turn off these individual modules.

Below are example commands to generate reports using these functionalities.

## A test run for human

```
runReport()
```
## A custom run for human
```
runReport( queryFilePath = 'input.BED',
            gffFilePath = 'annotation.gtf')
```
## To turn off certain modules of the report
``` 
runReport( queryFilePath = 'input.BED',
            gffFilePath = 'annotation.gtf',
            motifAnalysis = FALSE,
            goAnalysis = FALSE )
```
## To run the pipeline for species other than human

You can run RCAS for any of the genome versions available in the BSgenome package.  
See `BSgenome::available.genomes`. 

```
runReport( queryFilePath = 'input.mm9.BED',
            gffFilePath = 'annotation.mm9.gtf',
            genomeVersion = 'mm9' )
```

## To turn off verbose output and progress bars
```
runReport(quiet = TRUE)
```

## Printing raw data generated by the runReport function
One may be interested in printing the raw data used to make the plots and tables
in the HTML report output of runReport function. Such tables could be used for
meta-analysis of multiple analysis results. In order to activate this function,
printProcessedTables argument must be set to TRUE.
```
runReport(printProcessedTables = TRUE)
```

# Acknowledgements

RCAS is developed in the group of 
[Altuna Akalin](http://bioinformatics.mdc-berlin.de/team.html#altuna-akalin-phd)
(head of the Scientific Bioinformatics Platform) by
[Bora Uyar](http://bioinformatics.mdc-berlin.de/team.html#bora-uyar-phd)
(Bioinformatics Scientist),
[Dilmurat Yusuf](http://bioinformatics.mdc-berlin.de/team.html#dilmurat-yusuf-phd)
(Bioinformatics Scientist) and 
[Ricardo Wurmus](http://bioinformatics.mdc-berlin.de/team.html#ricardo-wurmus)
(System Administrator) at the Berlin Institute of Medical Systems Biology
([BIMSB](https://www.mdc-berlin.de/13800178/en/bimsb))
at the Max-Delbrueck-Center for Molecular Medicine
([MDC](https://www.mdc-berlin.de)) in Berlin.

RCAS is developed as a bioinformatics service as part of
the [RNA Bioinformatics Center](http://www.denbi.de/index.php/rbc),
which is one of the eight centers of
the German Network for Bioinformatics Infrastructure
([de.NBI](http://www.denbi.de/)).  

