---
title: "Enrichment Correlation"
output: html_document
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=FALSE, include=FALSE)
```

# Load paths and libraries
```{r include=FALSE}
source("shared_functions.R")
source("combined_config.R")

# source("run_config.R")
# library(foreach)
# library(dplyr)
# library(tidyr)
# library(magrittr)
# library(rtracklayer)
# library(ggplot2)
# library(gridExtra)
# library(readr)
# library(stringr)
# library(Rsamtools)
# library(DESeq2)
# library(tibble)
```

## Load STAR Count Data
```{r}
starout_2019_dir
```


```{r}
list.files(starout_2019_dir)
           
           
           
           , pattern = paste0(count_suffix,"$"), full.names = TRUE)



countfiles = list.files(starout_2019_dir, pattern = paste0(count_suffix,"$"), full.names = TRUE)
countfiles
```

### Load STAR Count tables into a single dataframe
Loading code borrowed from https://gitlab.oit.duke.edu/HTS2018/HTS2018-notebooks/blob/master/pilot/01_Read_Counts.ipynb
```{r}
countfile_paths %>%
  loadStarCounts()  ->
  genecounts



genecounts[1:6,1:6]
```

### Extract library specific label
```{r}
genecounts %>%
  mutate(Label=str_extract(expid, "^\\d+_[A-Z]+_[A-Z]")) %>%
  dplyr::select(Label, everything()) %>%
  dplyr::select(-expid) ->
  genecounts_with_label
genecounts_with_label[1:8,1:6]
```

### Combine per library counts across lanes
```{r}
genecounts_with_label %>%
  group_by(Label) %>%
  summarise_all(funs(sum)) ->
  genecounts_lanes_combined

# Add metadata 
read_tsv(metadata.file) %>%
  dplyr::select(Label, enrichment_method, RNA_sample_num) ->
  metadata.df

genecounts_lanes_combined %<>%
  left_join(metadata.df, by="Label") %>%
  dplyr::select(library=Label, enrichment_method, RNA_sample_num, everything())
genecounts_lanes_combined[,1:6]
```

## Compare library preps

### Drop rRNA genes and genes with rRNA homology

#### Find all genes that are mapped by rRNA reads
##### Load GTF
```{r}
import(gtf_with_mito_rrna.file, format="gtf") %>%
  subset(type=="exon") ->
  h99_granges
```


##### Find all genes that overlap with rRNA reads
https://bioinformatics.stackexchange.com/questions/874/intersection-of-two-genomic-ranges-to-keep-metadata
```{r}
scanBam(oligo_bam.file) %>%
  as.data.frame %>%
  mutate(end=pos+qwidth) %>%
  dplyr::select(chr=rname, start=pos, end, strand) %>%
  makeGRangesFromDataFrame(ignore.strand = TRUE) %>%
  invertStrand -> # we invert the strand since the oligos are antisense
  oligo_bam.gr

# %>% rev
  # oligo_gene_counts

o = findOverlaps(h99_granges, oligo_bam.gr)
grl1 = split(h99_granges[queryHits(o)], 1:length(o)) # You can't mendoapply on a GRanges object
grl2 = split(oligo_bam.gr[subjectHits(o)], 1:length(o))
foo = function(x, y) {
  rv = x
  start(rv) = max(start(x), start(y))
  end(rv) = min(end(x), end(y))
  return(rv)
}

unlist(mendoapply(foo, grl1, y=grl2)) %>%
  as.data.frame %>%
  pull(gene_id) %>%
  unique ->
  all_rrna_homologs

all_rrna_homologs
```

##### Find STRAND-SPECIFIC genes that overlap with rRNA reads
```{r}
scanBam(oligo_bam.file) %>%
  as.data.frame %>%
  mutate(end=pos+qwidth) %>%
  dplyr::select(chr=rname, start=pos, end, strand) %>%
  makeGRangesFromDataFrame %>%
  invertStrand -> # we invert the strand since the oligos are antisense
  oligo_bam.gr

# %>% rev
  # oligo_gene_counts

o = findOverlaps(h99_granges, oligo_bam.gr)
grl1 = split(h99_granges[queryHits(o)], 1:length(o)) # You can't mendoapply on a GRanges object
grl2 = split(oligo_bam.gr[subjectHits(o)], 1:length(o))
foo = function(x, y) {
  rv = x
  start(rv) = max(start(x), start(y))
  end(rv) = min(end(x), end(y))
  return(rv)
}

unlist(mendoapply(foo, grl1, y=grl2)) %>%
  as.data.frame %>%
  pull(gene_id) %>%
  unique ->
  strand_specific_rrna_homologs

strand_specific_rrna_homologs
```

#### Drop genes from count table

```{r}
# genes_to_drop = intersect(colnames(genecounts_lanes_combined), 
#                           strand_specific_rrna_homologs)
genecounts_lanes_combined %>%
  dplyr::select(strand_specific_rrna_homologs) ->
  check_select

check_select[,1:6]

genecounts_lanes_combined %>%
  dplyr::select(-strand_specific_rrna_homologs) ->
  genecounts_lanes_combined_no_rrna

genecounts_lanes_combined_no_rrna[,1:6]

dim(check_select)
dim(genecounts_lanes_combined)
dim(genecounts_lanes_combined_no_rrna)

dplyr::select(genecounts_lanes_combined,small_MTrRNA)
# dplyr::select(genecounts_lanes_combined_no_rrna,small_MTrRNA)

```


### Generate scatterplots for MA vs TOT and RZ vs TOT
```{r}
genecounts_lanes_combined_no_rrna[,1:6]

```

```{r}

cur_sample_num = 2

genecounts_lanes_combined_no_rrna %>%
  filter(RNA_sample_num == cur_sample_num) %>%
  dplyr::select(-c(library,RNA_sample_num)) %>%
  gather(gene_id, reads, 2:ncol(.)) %>%
  spread(enrichment_method,reads) ->
  cur_sample.tidy
  
title_text = paste0("Sample #", cur_sample_num, collapse=" ")

cur_sample.tidy %>%
  summarise(max(MA), max(RZ)) %>%
  max -> ymax

tot_vs_ma = ggplot(cur_sample.tidy, aes(TOT, MA)) + 
  geom_point(alpha = 1/2, color="red", size=1) +
  ggtitle(title_text) +
  theme_bw() +
  ylim(0, ymax)
tot_vs_rz = ggplot(cur_sample.tidy, aes(TOT, RZ)) + 
  geom_point(alpha = 1/2, color="red", size=1) +
  ggtitle(title_text) +
  theme_bw() +
  ylim(0, ymax)

grid.arrange(tot_vs_ma, tot_vs_rz, nrow = 1)

#------------------------------------------------------------------
ma_cor = paste("MA R =", 
               cor(cur_sample.tidy[["TOT"]],
                   cur_sample.tidy[["MA"]]) %>%
                 round(2))

rz_cor = paste("RZ R =", 
               cor(cur_sample.tidy[["TOT"]],
                   cur_sample.tidy[["RZ"]]) %>%
                 round(2))
ma_cor
rz_cor
cat(paste(ma_cor, rz_cor, sep="\n"))
#------------------------------------------------------------------
ggplot(cur_sample.tidy, aes(TOT, y=value, color=variable, alpha = 1/2)) + 
  geom_point(aes(y=MA, color="MA"), size=1) +
  geom_point(aes(y=RZ, color="RZ"), size=1) +
  ggtitle(title_text) +
  theme_bw() +
  annotate("text", x=0, y = ymax*0.95, 
           label = paste(ma_cor, rz_cor, sep="\n"),
           hjust="inward",
           just="inward", 
           color="red")

#------------------------------------------------------------------

ggplot(cur_sample.tidy, alpha = 1/2) +
  geom_point(aes(TOT, MA), color="red", size=1) + 
  geom_smooth(aes(TOT, MA), method=lm, se=FALSE, color="red") +
  geom_point(aes(TOT, RZ), color="blue", size=1) +
  geom_smooth(aes(TOT, RZ), method=lm, se=FALSE, color="blue") +
  ggtitle(title_text) +
  theme_bw() 


```

```{r include=TRUE}
makeScatterPlot = function(genecounts_lanes_combined_no_rrna, cur_sample_num) {
  
genecounts_lanes_combined_no_rrna %>%
  filter(RNA_sample_num == cur_sample_num) %>%
  dplyr::select(-c(library,RNA_sample_num)) %>%
  gather(gene_id, reads, 2:ncol(.)) %>%
  spread(enrichment_method,reads) ->
  cur_sample.tidy
  
title_text = paste0("Sample #", cur_sample_num, collapse=" ")

cur_sample.tidy %>%
  summarise(max(MA), max(RZ)) %>%
  max -> ymax

#------------------------------------------------------------------
ma_cor = paste("MA R =", 
               cor(cur_sample.tidy[["TOT"]],
                   cur_sample.tidy[["MA"]]) %>%
                 round(2))

rz_cor = paste("RZ R =", 
               cor(cur_sample.tidy[["TOT"]],
                   cur_sample.tidy[["RZ"]]) %>%
                 round(2))
#------------------------------------------------------------------
ggplot(cur_sample.tidy, aes(TOT, y=value, color=variable, alpha = 1/2)) + 
  geom_point(aes(y=MA, color="MA"), size=1) +
  geom_point(aes(y=RZ, color="RZ"), size=1) +
  ggtitle(title_text) +
  theme_bw() +
  annotate("text", x=0, y = ymax, 
           label = paste(ma_cor, rz_cor, sep="\n"),
           hjust="inward",
           vjust="inward", 
           color="black")
}
for (cur_sample in 2:4) {
  print(makeScatterPlot(genecounts_lanes_combined_no_rrna, cur_sample))
}
```


```{r}
# Trying to generalize plotting 
require(dplyr)
set.seed(123)
xx = data.frame(group = rep(1:4, 100), a = rnorm(400) , b = rnorm(400))

xx
gp = group_by(xx, group)
dplyr::summarize(gp, cor(a, b))

# https://www.sixhat.net/how-to-plot-multpile-data-series-with-ggplot.html
```

# Analysis with Normalized Counts


## Load Data in DESeq2

### Make Count Matrix for DESeq
```{r}
genecounts %>%
  select(-strand_specific_rrna_homologs) %>% # drop rRNA and rRNA homologs
  column_to_rownames(var="expid") %>%
  as.matrix %>%
  t ->
  genecounts_for_desesq
genecounts_for_desesq[1:6,1:6]
```

### Make Metadata dataframe for DESeq
```{r}
# read_tsv(metadata.file) %>%
#   dplyr::select(Label, enrichment_method, RNA_sample_num) ->
#   metadata.df

genecounts %>%
  select(expid) %>%
  mutate(library=str_extract(expid, "^\\d+_[A-Z]+_[A-Z]")) %>%
  left_join(metadata.df, by=c("library"="Label")) %>%
  mutate(enrichment_method=factor(enrichment_method, 
                                  levels = c("TOT", "MA", "RZ"))) %>%
  column_to_rownames(var="expid") ->
  sample_df
head(sample_df)
```

### Make DESeq object
#### Confirm that sample metadata and count columns are in the same order
```{r}
all(rownames(sample_df) == colnames(genecounts_for_desesq))
```

#### Make DESeqDataSet
```{r}
DESeqDataSetFromMatrix(countData = genecounts_for_desesq,
                       colData = sample_df,
                       design = ~ enrichment_method) ->
  dds
colData(dds)

# Collapse Replicates
dds %>% collapseReplicates(object=., 
                           groupby=colData(.) %>% 
                             as.tibble %>%
                             pull(library)) ->
  dds
dds
```

```{r}
dds <- DESeq(dds)
counts(dds,normalized=TRUE)
results(dds, contrast = c("enrichment_method", "MA", "MA"))
```
```{r}
dds %>%
  counts(normalized=TRUE) ->
  norm_counts

norm_counts %>%
  as_tibble(rownames = "geneid") %>%
  mutate(ma_mean = rowMeans(select(., contains("MA"))),
         rz_mean = rowMeans(select(., contains("RZ"))),
         tot_mean = rowMeans(select(., contains("TOT")))) ->
  norm_counts_wmeans


norm_counts_wmeans %>%
  ggplot(aes(x=tot_mean, y=rz_mean)) +
  geom_point()

norm_counts_wmeans %>%
  {cor(pull(., tot_mean), pull(., rz_mean))} ->
  rz_cor

norm_counts_wmeans %>%
  {cor(pull(., tot_mean), pull(., ma_mean))} ->
  ma_cor
```

```{r}

norm_counts_wmeans %>%
  summarise(max(ma_mean), max(rz_mean)) %>%
  max -> ymax


norm_counts_wmeans %>%
  ggplot(aes(x=tot_mean, y=value, color=variable, alpha = 1/2)) + 
  geom_point(aes(y=ma_mean, color="ma_mean"), size=1) +
  geom_point(aes(y=rz_mean, color="rz_mean"), size=1) +
  theme_classic() +
  annotate("text", x=0, y = ymax, 
           label = paste("MA R=", round(ma_cor,2), "\nRZ R=", round(rz_cor,2)),
           hjust="inward",
           vjust="inward", 
           color="black")
```






```{r}
sapply(levels(dds$enrichment_method), 
       function(lvl) rowMeans( counts(dds,normalized=TRUE)[,dds$enrichment_method == lvl, drop=F] ) )
```


```{r}
dds %>%
  counts(normalized=TRUE) %>%
  rowMeans

```



## Work


```{r}
rld <- rlog(dds, blind=TRUE)
rld
```

```{r}
counts(dds,normalized=TRUE)
```

counts(dds, normalized=TRUE)

Heatmap of the count matrix


```{r}
rlog(dds, blind=TRUE) %>%
  counts
  
```




# Add metadata 
